---
title: "Differential expression analysis"
author: "André F. B. Menezes, Beatrice Tropea"
date: "`r format(Sys.time(), '%B %d, %Y')`"
toc: true
number-sections: true
format:
  html:
    html-math-method: katex
    code-tools: true
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
---


# Introduction {#sec-intro}

In this report we will perform a differential expression analysis of proteins with a biological relevance exctracted . 

```{r}
fts <- readRDS(file = "../data/fts_processed.rds")
fts[["proteins"]]
rowData(fts[["proteins"]])
se <- fts[["proteins"]]
mean(assay(se, "log2_normalized"))
# Filters out the variable proteins
chosen_proteins <- rownames(se)[rowData(se)$bio > 0]
dim(se)
length(chosen_proteins)
```
## Perform DE analysis using limma package
```{r}
se_curr <- se[chosen_proteins, ]
assay(se)
g <- factor(colData(se_curr)$group)
g <- relevel(g, ref = "Control")

design_matrix <- model.matrix(~ g)
colnames(design_matrix) <- levels(g)

# Fitting linear model --------------------------------------------------------
fit <- limma::lmFit(object = assay(se_curr, "log2_normalized"), design_matrix)
fit <- limma::eBayes(fit)

ampicillin_control <- limma::topTable(fit, coef = "Ampicillin", number = Inf,
                               sort.by = "none", confint = TRUE)
cefotaxime_control <- limma::topTable(fit, coef = "Cefotaxime", number = Inf,
                               sort.by = "none", confint = TRUE)
impipenem_control <- limma::topTable(fit, coef = "Impipenem", number = Inf,
                              sort.by = "none", confint = TRUE)
ciprofloxacin_control <- limma::topTable(fit, coef = "Ciprofloxacin", number = Inf,
                                  sort.by = "none", confint = TRUE)

```

```{r}
suppressMessages(library(QFeatures))
suppressMessages(library(dplyr))
library(tidyr)
library(limma)

# Selecting necessary columns and renaming them
amp <- ampicillin_control |>
  select(logFC, P.Value, adj.P.Val) |>
  rename("logFC" = logFC, "P_value" = P.Value, 
         "FDR" = adj.P.Val)
Protein <- row.names(amp)
amp <- cbind(Protein, 'Treatment'="Ampicillin", amp)

cef <- cefotaxime_control |>
  select(logFC, P.Value, adj.P.Val) |>
  rename("logFC" = logFC, "P_value" = P.Value, 
         "FDR" = adj.P.Val)
Protein <- row.names(cef)
cef <- cbind(Protein, 'Treatment'="Cefotaxime", cef)

imp <- impipenem_control |>
  select(logFC, P.Value, adj.P.Val) |>
  rename("logFC" = logFC, "P_value" = P.Value, 
         "FDR" = adj.P.Val)
Protein <- row.names(imp)
imp <- cbind(Protein, 'Treatment'="Impipenem", imp)

cip <- ciprofloxacin_control |>
  select(logFC, P.Value, adj.P.Val) |>
  rename("logFC" = logFC, "P_value" = P.Value, 
         "FDR" = adj.P.Val)
Protein <- row.names(cip)
cip <- cbind(Protein, 'Treatment'="Ciprofloxacin", cip)

# Creating data frame
limma_data <- rbind(amp, cef, imp, cip)

```

```{r}
# exporting data
library(openxlsx)
write.xlsx(
  x = limma_data,
  file = file.path("./", "Qfeatures_Limma.xlsx"),
  rowNames = FALSE)
```

# VolcanoPlot

```{R}
library(ggplot2)
library(ggrepel)
#Ampicillin
data_summarized <- limma_data |>
  filter(Treatment %in% c("Ampicillin")) |>
  select(Protein, logFC, P_value)

  dfa1 <-  data_summarized %>%
  filter((logFC) >= (0.5) & (P_value) <= (0.05)) %>%
  arrange(desc(logFC)) %>%
  head(10)

dfa2 <-  data_summarized %>%
  filter(((logFC) <= -(0.5)) & (P_value) <= (0.05)) %>%
  arrange(logFC) %>%
  head(10)

dfa <- rbind(dfa1, dfa2[ ,])

p1 <- ggplot(data_summarized, aes((logFC), -log(P_value,10)), options(ggrepel.max.overlaps = Inf)) + # -log10 conversion
  geom_point(size = 0.005, col="gray") +
  xlab(expression("log"[2]*"FC")) + 
  ylab(expression("-log"[10]*"pvalue")) +
  ggtitle("Control Vs Ampicillin") +
  geom_vline(xintercept=c(-0.5, 0.5), linetype = "dashed", col="black") +
  geom_hline(yintercept=-log10(0.05), linetype = "dashed", col="black") +
  geom_point(data = dfa,
             shape = 21,
             size = 0.5,
             fill = "red",
             colour = "red") +
  geom_text_repel(data=dfa,
                  aes(x=logFC,
                      y=-log(P_value,10),
                      label=Protein),
                  size=1.5,
                  color="black") +
  theme_bw()

save_plot(file.path("./", "QVolcanoPlot_Ampicillin.png"), p1)
```

```{R}
#Imipenem

data_summarized2 <- limma_data |>
  filter(Treatment %in% c("Impipenem")) |>
  select(Protein, logFC, P_value)

  dfi1 <-  data_summarized2 %>%
  filter((logFC) >= (0.5) & (P_value) <= (0.05)) %>%
  arrange(desc(logFC)) %>%
  head(10)

dfi2 <-  data_summarized2 %>%
  filter(((logFC) <= -(0.5)) & (P_value) <= (0.05)) %>%
  arrange(logFC) %>%
  head(10)

dfi <- rbind(dfi1, dfi2[ ,])
  
p2 <- ggplot(data_summarized2, aes((logFC), -log(P_value,10)), options(ggrepel.max.overlaps = Inf)) + # -log10 conversion
  geom_point(size = 0.005, col="gray") +
  xlab(expression("log"[2]*"FC")) +
  ylab(expression("-log"[10]*"pvalue")) +
  ggtitle("Control Vs Imipenem") +
  geom_vline(xintercept=c(-0.5, 0.5), linetype = "dashed", col="black") +
  geom_hline(yintercept=-log10(0.05), linetype = "dashed", col="black") +
  geom_point(data = dfi,
             shape = 21,
             size = 0.5,
             fill = "red",
             colour = "red") +
  geom_text_repel(data=dfi,
                  aes(x=logFC,
                      y=-log(P_value,10),
                      label=Protein),
                  size=1.5,
                  color="black") +
  theme_bw()

save_plot(file.path("./", "QVolcanoPlot_Imipenem.png"), p2)
```
```{R}
#Cefotaxime

data_summarized3 <- limma_data |>
  filter(Treatment %in% c("Cefotaxime")) |>
  select(Protein, logFC, P_value)

  dfce1 <-  data_summarized3 %>%
  filter((logFC) >= (0.5) & (P_value) <= (0.05)) %>%
  arrange(desc(logFC)) %>%
  head(10)

dfce2 <-  data_summarized3 %>%
  filter(((logFC) <= -(0.5)) & (P_value) <= (0.05)) %>%
  arrange(logFC) %>%
  head(10)

dfce <- rbind(dfce1, dfce2[ ,])

p3 <- ggplot(data_summarized3, aes((logFC), -log(P_value,10)), options(ggrepel.max.overlaps = Inf)) + # -log10 conversion
  geom_point(size = 0.005, col="gray") +
  xlab(expression("log"[2]*"FC")) +
  ylab(expression("-log"[10]*"pvalue")) +
  ggtitle("Control Vs Cefotaxime") +
  geom_vline(xintercept=c(-0.5, 0.5), linetype = "dashed", col="black") +
  geom_hline(yintercept=-log10(0.05), linetype = "dashed", col="black") +
  geom_point(data = dfce,
             shape = 21,
             size = 0.5,
             fill = "red",
             colour = "red") +
  geom_text_repel(data=dfce,
                  aes(x=logFC,
                      y=-log(P_value,10),
                      label=Protein),
                  size=1.5,
                  color="black") +
  theme_bw()

save_plot(file.path("./", "QVolcanoPlot_Cefotaxime.png"), p3)
```
          
```{R}
#Ciprofloxacin
data_summarized4 <- limma_data |>
  filter(Treatment %in% c("Ciprofloxacin")) |>
  select(Protein, logFC, P_value)

  dfci1 <-  data_summarized %>%
  filter((logFC) >= (0.5) & (P_value) <= (0.05)) %>%
  arrange(desc(logFC)) %>%
  head(10)

dfci2 <-  data_summarized %>%
  filter(((logFC) <= -(0.5)) & (P_value) <= (0.05)) %>%
  arrange(logFC) %>%
  head(10)

dfci <- rbind(dfci1, dfci2[ ,])
  
p4 <- ggplot(data_summarized, aes((logFC), -log(P_value,10)), options(ggrepel.max.overlaps = Inf)) + # -log10 conversion
  geom_point(size = 0.005, col="gray") +
  xlab(expression("log"[2]*"FC")) +
  ylab(expression("-log"[10]*"pvalue")) +
  ggtitle("Control Vs Ciprofloxacin") +
  geom_vline(xintercept=c(-0.5, 0.5), linetype = "dashed", col="black") +
  geom_hline(yintercept=-log10(0.05), linetype = "dashed", col="black") +
  geom_point(data = dfci,
             shape = 21,
             size = 0.5,
             fill = "red",
             colour = "red") +
  geom_text_repel(data=dfci,
                  aes(x=logFC,
                      y=-log(P_value,10),
                      label=Protein),
                  size=1.5,
                  color="black") +
  theme_bw()

save_plot(file.path("./", "QVolcanoPlot_Ciprofloxacin.png"), p4)
```

# Comparison with results showed in the paper '..'

```{R}
library(readxl)
library(writexl)
dataFiona <- read_xlsx(file.path("./", "all_results.xlsx"))

# ----USAGE: to perform the selected analysis -> corresponding function(treatment)
#   in which treatment= "Ampicillin"|"Impipenem"|"Cefotaxime"|"Ciprofloxacin"------------

### perform traditional analysis with |logFC| > 0.5 & pvalue < 0.05 (sorting by pvalue)
P <- function(treatment, data, name="no") {
  data_filtered <- data |>
    filter(Treatment %in% c(treatment)) |>
    select(c(Protein, logFC, P_value))

df1 <- data_filtered %>%
  filter(((logFC) >= (0.5)) & ((P_value) <= (0.05))) %>%
  arrange(P_value)
  if (name=="yes") {df1 <- df1 %>% head(10)}

df2 <-  data_filtered %>%
  filter(((logFC) <= -(0.5)) & ((P_value) <= (0.05))) %>%
  arrange(P_value)
  if (name=="yes") {df2 <- df2 %>% head(10)}

df <- rbind(df1, df2[ ,])
return(df)}

### perform traditional analysis with |logFC| > 0.5 & pvalue < 0.05 (sorting by logFC - paper)
Original <- function(treatment, data, name="no") {
  data_filtered <- data |>
    filter(Treatment %in% c(treatment)) |>
    select(c(Protein, logFC, P_value))
  
  df1 <- data_filtered %>%
    filter(((logFC) >= (0.5)) & ((P_value) <= (0.05))) %>%
    arrange(desc(logFC))
  if (name=="yes") {df1 <- df1 %>% head(10)}
  
  df2 <-  data_filtered %>%
    filter(((logFC) <= -(0.5)) & ((P_value) <= (0.05))) %>%
    arrange(logFC)
  if (name=="yes") {df2 <- df2 %>% head(10)}
  
  df <- rbind(df1, df2[ ,])
  return(df)}

### perform more restrictive analysis with |logFC| > 0.5 & pvalue < 0.005
newP <- function(treatment, data, name="no") {
  data_filtered <- data |>
    filter(Treatment %in% c(treatment)) |>
    select(c(Protein, logFC, P_value))
  
df1 <- data_filtered  %>%
  filter((logFC) >= (0.5) & (P_value) <= (0.005)) %>%
  arrange(desc(logFC))
if (name=="yes") {df1 <- df1 %>% head(10)}

df2 <-  data_filtered %>%
  filter(((logFC) <= -(0.5)) & (P_value) <= (0.005)) %>%
  arrange(logFC) 
if (name=="yes") {df2 <- df2 %>% head(10)}

df <- rbind(df1, df2[ ,])
return(df)}

### perform analysis with |logFC| > 0.5 & p value adjustment (FDR < 0.1)
FDR <- function(treatment, data, name="no") {
  data_filtered <- data |>
    filter(Treatment %in% c(treatment)) |>
    select(c(Protein, logFC, FDR))
  
df1 <-  data_filtered %>%
  filter((logFC) >= (0.5) & (FDR) <= (0.05)) %>%
  arrange(FDR)
if (name=="yes") {df1 <- df1 %>% head(10)}

df2 <-  data_filtered %>%
  filter(((logFC) <= -(0.5)) & (FDR) <= (0.1)) %>%
  arrange(FDR)
if (name=="yes") {df2 <- df2 %>% head(10)}

df <- rbind(df1, df2[ ,])
return(df)}
```

## Protein Differential Expressed in each Treatment based on standard cutoff of Pvalue <= 0.05 & |logFC| >= 0.5 - Bar Chart

```{R}
# Fiona Rules----------------------------------------------------------------------------------

Ampicillin <- as.numeric(tibble(nrow(P("Ampicillin", dataFiona)), nrow(P("Ampicillin", limma_data))))
Approaches <- c("FionaData", "newData")
Ampicillin_proteinDE_P <- cbind(Approaches, Ampicillin)
Ampicillin_proteinDE_P <- as.data.frame(Ampicillin_proteinDE_P)

Impipenem <- as.numeric(tibble(nrow(P("Impipenem", dataFiona)), nrow(P("Impipenem", limma_data))))
Approaches <- c("FionaData", "newData")
Impipenem_proteinDE_P <- cbind(Approaches, Impipenem)
Impipenem_proteinDE_P <- as.data.frame(Impipenem_proteinDE_P)

Cefotaxime <- as.numeric(tibble(nrow(P("Cefotaxime", dataFiona)), nrow(P("Cefotaxime", limma_data))))
Approaches <- c("FionaData", "newData")
Cefotaxime_proteinDE_P <- cbind(Approaches, Cefotaxime)
Cefotaxime_proteinDE_P <- as.data.frame(Cefotaxime_proteinDE_P)

Ciprofloxacin <- as.numeric(tibble(nrow(P("Ciprofloxacin", dataFiona)), nrow(P("Ciprofloxacin", limma_data))))
Approaches <- c("FionaData", "newData")
Ciprofloxacin_proteinDE_P <- cbind(Approaches, Ciprofloxacin)
Ciprofloxacin_proteinDE_P <- as.data.frame(Ciprofloxacin_proteinDE_P)

sumupCC <- right_join(Cefotaxime_proteinDE_P , Ciprofloxacin_proteinDE_P, by="Approaches")
sumupAI <- right_join(Ampicillin_proteinDE_P, Impipenem_proteinDE_P, by="Approaches")
sumup <- right_join(sumupAI, sumupCC, by="Approaches")
graph <- sumup %>% pivot_longer(!Approaches, names_to = 'Treatment')

tot1 <- graph %>%
  ggplot(aes(x = Approaches, y = as.numeric(value), 
             fill = Treatment)) +
  ylab(expression("# Protein DE")) +
  ggtitle("Comparison Dataset using cutoff of pvalue <= 0.05") +
  geom_bar(position = position_dodge(), stat="identity") +
  geom_text(aes(label = value), position=position_dodge(width=0.9), size=4)
tot1
save_plot(file.path("./", "ProteinDE_BarChart_StandardP.png"), tot1)
```
## Protein Differential Expressed in each Treatment based on cutoff of Pvalue <= 0.005 & |logFC| >= 0.5 - Bar Chart

```{R}

Ampicillin <- as.numeric(tibble(nrow(newP("Ampicillin", dataFiona)), nrow(newP("Ampicillin", limma_data))))
Approaches <- c("FionaData", "newData")
Ampicillin_proteinDE_newP <- cbind(Approaches, Ampicillin)
Ampicillin_proteinDE_newP <- as.data.frame(Ampicillin_proteinDE_newP)

Impipenem<- as.numeric(tibble(nrow(newP("Impipenem", dataFiona)), nrow(newP("Impipenem", limma_data))))
Approaches <- c("FionaData", "newData")
Impipenem_proteinDE_newP <- cbind(Approaches, Impipenem)
Impipenem_proteinDE_newP <- as.data.frame(Impipenem_proteinDE_newP)

Cefotaxime <- as.numeric(tibble(nrow(newP("Cefotaxime", dataFiona)), nrow(newP("Cefotaxime", limma_data))))
Approaches <- c("FionaData", "newData")
Cefotaxime_proteinDE_newP <- cbind(Approaches, Cefotaxime)
Cefotaxime_proteinDE_newP <- as.data.frame(Cefotaxime_proteinDE_newP)

Ciprofloxacin <- as.numeric(tibble(nrow(newP("Ciprofloxacin", dataFiona)), nrow(newP("Ciprofloxacin", limma_data))))
Approaches <- c("FionaData", "newData")
Ciprofloxacin_proteinDE_newP <- cbind(Approaches, Ciprofloxacin)
Ciprofloxacin_proteinDE_newP <- as.data.frame(Ciprofloxacin_proteinDE_newP)
# Summary Bar Chart of how many differential expressed protein each approaches identify for each treatment

## preparing table of data
sumupCC <- right_join(Cefotaxime_proteinDE_newP , Ciprofloxacin_proteinDE_newP, by="Approaches")
sumupAI <- right_join(Ampicillin_proteinDE_newP, Impipenem_proteinDE_newP, by="Approaches")
sumup <- right_join(sumupAI, sumupCC, by="Approaches")
graph <- sumup %>% pivot_longer(!Approaches, names_to = 'Treatment')

## plotting
tot2 <- graph %>%
  ggplot(aes(x = Approaches, y = as.numeric(value), 
             fill = Treatment)) +
  ylab(expression("# Protein DE")) +
  ggtitle("Comparison Dataset using cutoff of pvalue <= 0.005") +
  geom_bar(position = position_dodge(), stat="identity") +
  geom_text(aes(label = value), position=position_dodge(width=0.9), size=4)
tot2
save_plot(file.path("./", "ProteinDE_BarChart_newP.png"), tot2)
```
```{R}
Original("Ciprofloxacin", dataFiona, "yes")
Original("Ciprofloxacin", limma_data, "yes")
```